Certified Knowledge Compilation \\ with Application to Verified Model Counting

### 2023-03-18 doc/mc-paper.tex
Minor changes

Many important properties of Boolean formulas can be
expressed as ring evaluation problems.  The
{\em model counting} problem for formula $\phi$ requires determining $|\modelset(\phi)|$.
It can be cast as a ring evaluation problem by having $\radd$ and
$\rmul$ be addition and multiplication over rational numbers and using
label $w(x) = 1/2$ for every variable $x$.
Ring evaluation of formula $\phi$ gives the {\em density} of
the formula, i.e., the fraction of all possible total assignments that are
models.  For $n = |\varset|$, scaling the density by $2^n$
yields the number of models.

@@

POGs are inspired by the d-DNNF graphs devised by Darwiche for
representing Boolean functions \cite{darwiche:jair:2002}.
POGs generalize d-DNNF in two ways:
\begin{itemize}
\item They allow negation of arbitrary nodes in the graph, not just
  variables.  As we have seen, ring evaluation can readily handle negation.
  %, and so we allow it in the interest of generality.
%%  Negation could be useful in some contexts, such as when
%%  converting binary decision diagrams (BDDs) having complemented
%%  edges~\cite{brace-dac-1990,minato-dac-1990} into POGs.

@@

A proof in our framework takes the form of a sequence of clause addition
and deletion steps, with each step preserving the solutions to
the original formula.  The status of the proof at any step is represented as
a set of {\em active} clauses $\phi$, i.e., those that
have been added but not yet deleted.  Our framework is based
on {\em extended} resolution~\cite{Tseitin:1983}, meaning that proof
steps can introduce new {\em extension variables} encoding Boolean formulas over input and prior extension variables.
Let $Z$
denote the set of extension variables occuring in formula $\phi$.
Starting with input formula $\inputformula$,
the proof must maintain the invariant that
$\inputformula \ifandonlyif \exists Z\,\phi$.

@@

Compared to the proofs of unsatisfiability generated by SAT solvers,
it can be seen that ours has important differences.  Most
significantly, we want our proofs to maintain preserve the set of solutions at
each step, whereas an unsatisfiability proof need only guarantee that
a proof step does not makes a satisfiable set of clauses become
unsatisfiable.  In particular, our proofs must justify clause deletions as well as clause additions.

@@

A CRAT file provides both a declaration of a POG, as well as an easily checkable
proof that a Boolean formula, given in conjunctive normal
form, is logically equivalent to the POG, thus enabling certified
knowledge compilation.
%% The proof system is based on extended
%% resolution~\cite{Tseitin:1983} with reverse unit propagation
%% (RUP)~\cite{goldberg,vangelder08_verifying_rup_proofs} as the core method for
%% proving that a set of clauses logically implies another clause.
%% Alternate version
%% knowledge compilation.  The proof system is based on extended
%% resolution~\cite{Tseitin:1983} with reverse unit propagation
%% (RUP)~\cite{goldberg,vangelder08_verifying_rup_proofs} as the core method for
%% proving that a set of clauses logically implies another clause.
%%
%% As an example, consider the formula $\varphi = (\obar{a} \lor b) \land (\obar{b} \lor c \lor \obar{e}) \land (\obar c \lor d)$
%% and a RUP step to derive the target clause $(\obar{a} \lor d \lor \obar{e})$ from the three clauses from $\varphi$.
%% A RUP proof would take the following form.  We start with the assignment that falsifies the target clause. This assignment
%% is extended by clauses in $\varphi$ that become unit and eventually the empty clause. In the illustration below, the
%% involved clauses (antecedents) are shown in the order that they became unit / falsified.
%%
%% \begin{center}
%%   \begin{tabular}{lcccc}
%%          & \makebox[15mm]{Target}    & \makebox[10mm]{} & \makebox[10mm]{Antecedents} & \makebox[10mm]{} \\
%%   Clause & $\obar{a} \lor d \lor \obar{e}$ & $\obar{c} \lor d$  & $\obar{e} \lor \obar{b} \lor c$   & $\obar{a} \lor b$ \\
%%   \midrule
%%   Units  & $a$, $\obar{d}$, $e$      &  $\obar{c}$             & $\obar{b}$                         & $\nil$ \\
%%   \end{tabular}
%% \end{center}
%%
%% RUP is an alternative formulation of resolution.  For target clause
%% $C$, it can be seen that applying resolution operations to the
%% antecedent clauses from right to left will derive a clause $C'$ such
%% that $C' \subseteq C$.  By {\em subsumption} %~\cite{philipp:lpar:2017},
%% we then have $C' \rightarrow C$.  Compared to listing each resolution
%% operation as a separate step, using RUP as the basic proof step makes
%% the proofs more compact.
%%
%% As is shown in Figure~\ref{fig:chain}
%% two programs are involved in producing a certified compilation of a formula:
%% \begin{itemize}
%% \item The {\em proof generator} produces a CRAT file that both defines the POG and gives a proof that the POG is logically equivalent to the input formula.
%% As the figure indicates, the generator starts with the output of another knowledge compiler, such as one generating a d-DNNF representation of the formula.
%% \item The {\em proof checker} ensures that all of the proof conditions are satisfied.
%% \end{itemize}
%% >>>>>>> 9c1793da858d00e5d4022d6d33c009848ba2083b

The CRAT format draws its inspiration from the LRAT~\cite{lrat} and
QRAT~\cite{heule:JAR2014} formats for unquantified and quantified Boolean formulas, respectively.
Key properties include:
\begin{itemize}
  \item
  The file contains declarations of $\pand$ and $\por$ operations to describe the POG.
  Declaring a node $\nodeu$
implicitly adds an {\em extension} variable $u$ and a set of {\em defining} clauses $\theta_{u}$
  encoding the product or sum operation.
  This is the only means for adding extension variables to the proof.
\item Boolean negation is supported implicitly by allowing the
  arguments of the $\por$ and $\pand$ operations to be literals and not just
  variables.
\item
  The file contains explicit clause addition steps.
  A clause can only be added if it is logically implied by the existing clauses.
  A sequence of clause identifiers must be listed as {\em hints} providing a RUP verification of the implication.
\item
  The file contains explicit clause deletion steps.
  A clause can only be deleted if it is logically implied by the remaining clauses.
  A sequence of clause identifiers must be listed as {\em hints} providing a RUP verification of the implication.
\item The checker must track the dependency set for every input and
  extension variable.  For each $\pand$ operation, the checker must ensure that the dependency sets for its arguments are disjoint.
  The associated extension variable has a dependency set equal to the union of those of its arguments.
\item Declaring a $\por$ operation requires a sequence of clauses
  providing a RUP proof that the arguments are mutually exclusive.
  Only binary $\por$ operations are allowed.  Generalizing to $k$ arguments would require
  $k\,(k-1)/2$ proofs of disjointness.
\end{itemize}

>>>>

Many important properties of Boolean formulas can be
expressed as ring evaluation problems.  The
{\em model counting} problem for formula $\phi$ requires determining $|\modelset(\phi)|$.
It can be cast as a ring evaluation problem by having $\radd$ and
$\rmul$ be addition and multiplication over rational numbers and using
label $w(x) = 1/2$ for every variable $x$.
Ring evaluation of formula $\phi$ gives the {\em density} of
the formula, i.e., the fraction of all possible total assignments that are
models.  For $n = |\varset|$, scaling the density by $2^n$
yields the number of models.  This formulation avoids the need for a ``smoothing'' operation,
in which redundant expressions are inserted into the formula~\cite{darwiche:jair:2002}.

@@

POGs are inspired by the d-DNNF graphs devised by Darwiche for
representing Boolean functions \cite{darwiche:jair:2002}.
POGs generalize d-DNNF in two ways:
\begin{itemize}
\item They allow negation of arbitrary nodes in the graph, not just
  variables.
  %, and so we allow it in the interest of generality.
%%  Negation could be useful in some contexts, such as when
%%  converting binary decision diagrams (BDDs) having complemented
%%  edges~\cite{brace-dac-1990,minato-dac-1990} into POGs.

@@

A proof in our framework takes the form of a sequence of clause addition
and deletion steps, with each step preserving the set of solutions to
the original formula.
The status of the proof at any step is represented as
a set of {\em active} clauses $\phi$, i.e., those that
have been added but not yet deleted.
Our framework is based
on {\em extended} resolution~\cite{Tseitin:1983}, where proof
steps can introduce new {\em extension variables} encoding Boolean formulas over input and prior extension variables.
Let $Z$
denote the set of extension variables occuring in formula $\phi$.
Starting with input formula $\inputformula$,
the proof must maintain the invariant that
$\inputformula \ifandonlyif \exists Z\,\phi$.

@@

Compared to the proofs of unsatisfiability generated by SAT solvers,
ours have important differences.  Most
significantly, each proof step must preserve the set of solutions with respect to the input variables.
By contrast, an unsatisfiability proof need only guarantee that
no proof step cause a satisfiable set of clauses to become
unsatisfiable.  As a consequence, our proofs must justify clause deletions as well as clause additions.

@@

%% The proof system is based on extended
%% resolution~\cite{Tseitin:1983} with reverse unit propagation
%% (RUP)~\cite{goldberg,vangelder08_verifying_rup_proofs} as the core method for
%% proving that a set of clauses logically implies another clause.
%% Alternate version
%% knowledge compilation.  The proof system is based on extended
%% resolution~\cite{Tseitin:1983} with reverse unit propagation
%% (RUP)~\cite{goldberg,vangelder08_verifying_rup_proofs} as the core method for
%% proving that a set of clauses logically implies another clause.
%%
%% As an example, consider the formula $\varphi = (\obar{a} \lor b) \land (\obar{b} \lor c \lor \obar{e}) \land (\obar c \lor d)$
%% and a RUP step to derive the target clause $(\obar{a} \lor d \lor \obar{e})$ from the three clauses from $\varphi$.
%% A RUP proof would take the following form.  We start with the assignment that falsifies the target clause. This assignment
%% is extended by clauses in $\varphi$ that become unit and eventually the empty clause. In the illustration below, the
%% involved clauses (antecedents) are shown in the order that they became unit / falsified.
%%
%% \begin{center}
%%   \begin{tabular}{lcccc}
%%          & \makebox[15mm]{Target}    & \makebox[10mm]{} & \makebox[10mm]{Antecedents} & \makebox[10mm]{} \\
%%   Clause & $\obar{a} \lor d \lor \obar{e}$ & $\obar{c} \lor d$  & $\obar{e} \lor \obar{b} \lor c$   & $\obar{a} \lor b$ \\
%%   \midrule
%%   Units  & $a$, $\obar{d}$, $e$      &  $\obar{c}$             & $\obar{b}$                         & $\nil$ \\
%%   \end{tabular}
%% \end{center}
%%
%% RUP is an alternative formulation of resolution.  For target clause
%% $C$, it can be seen that applying resolution operations to the
%% antecedent clauses from right to left will derive a clause $C'$ such
%% that $C' \subseteq C$.  By {\em subsumption} %~\cite{philipp:lpar:2017},
%% we then have $C' \rightarrow C$.  Compared to listing each resolution
%% operation as a separate step, using RUP as the basic proof step makes
%% the proofs more compact.
%%
%% As is shown in Figure~\ref{fig:chain}
%% two programs are involved in producing a certified compilation of a formula:
%% \begin{itemize}
%% \item The {\em proof generator} produces a CRAT file that both defines the POG and gives a proof that the POG is logically equivalent to the input formula.
%% As the figure indicates, the generator starts with the output of another knowledge compiler, such as one generating a d-DNNF representation of the formula.
%% \item The {\em proof checker} ensures that all of the proof conditions are satisfied.
%% \end{itemize}
%% >>>>>>> 9c1793da858d00e5d4022d6d33c009848ba2083b

A CRAT file provides both a declaration of a POG, as well as a checkable
proof that a Boolean formula, given in conjunctive normal
form, is logically equivalent to the POG\@.
The proof format draws its inspiration from the LRAT~\cite{lrat} and
QRAT~\cite{heule:JAR2014} formats for unquantified and quantified Boolean formulas, respectively.
Key properties include:
\begin{itemize}
  \item
  The file contains declarations of $\pand$ and $\por$ operations to describe the POG.
  Declaring a node $\nodeu$
implicitly adds an {\em extension} variable $u$ and a set of {\em defining} clauses $\theta_{u}$
  encoding the product or sum operation.
  This is the only means for adding extension variables to the proof.
\item Boolean negation is supported implicitly by allowing the
  arguments of the $\por$ and $\pand$ operations to be literals and not just
  variables.
\item
  The file contains explicit clause addition steps.
  A clause can only be added if it is logically implied by the existing clauses.
  A sequence of clause identifiers must be listed as {\em hints} providing a RUP verification of the implication.
\item
  The file contains explicit clause deletion steps.
  A clause can only be deleted if it is logically implied by the remaining clauses.
  A sequence of clause identifiers must be listed as {\em hints} providing a RUP verification of the implication.
\item The checker must track the dependency set for every input and
  extension variable.  For each $\pand$ operation, the checker must ensure that the dependency sets for its arguments are disjoint.
  The associated extension variable has a dependency set equal to the union of those of its arguments.
\item Declaring a $\por$ operation requires a sequence of clauses
  providing a RUP proof that the arguments are mutually exclusive.
  Only binary $\por$ operations are allowed.  Generalizing to $k$ arguments would require
  $k\,(k-1)/2$ proofs of disjointness.
\end{itemize}
